A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-07-16 text/discussion.Rmd
add discussion of Hahahan and Weinberg essays to discussion, update references

The basic problem in tumor research of this kind, namely that the search space,
~20,000 protein-coding genes is greater than the number of tumor samples
[@hornNetSigNetworkbasedDiscovery2018]. And that is not accounting for the fact
that ~30k non-coding genes and hundreds of thousands of non-coding regulatory
elements exist with oncogenic or tumor suppressor potential. Hence, we sought to
investigate cancer as a disease of basic cellular and biochemical pathways,
which we accomplished by translating gene-level mutations into pathway level
disruptions. Importantly, we selectively removed pathways that lacked generality
as cellular processes or that could be specifically assigned to differentiated
cell types, while excluding others.

We hypothesize that cancer is the result of dysfunction from a limited number of
basic cellular processes that are common to all living human cells, and thus all
tumors belong on a spectrum of dysfunction on this axis regardless of tissue of
origin. An alternative hypothesis is that every tumor belongs to one of a large
number of syndromes which are unique to each tissue of origin, that share some
mechanisms and superficial similarity in treatment strategies. The latter idea
has gained momentum in recent years since the publication of Encode, Roadmap and
TCGA consortium papers, perhaps owing to the intractable complexity of genomic
data in all its forms. The inescapable conclusion thus far is that all these
data point to tissue-of-origin as the most important driver of tumor behavior.
Our observations contrast with this view, and instead support an interpretation
in which all tumors manifest one of a limited number of metastable phenotypes
represented by our classes (**figure 1a-k**), with increased fitness regardless
of tissue of origin, which then converge on end stage cancer by accumulating
complementary mutations to the other classes (**figure 1l** and **Table 2**).

We find that most tumors segregate into one of 12 biologically distinct
clusters, regardless of the cancer type (**figure 1h** and **supplementary
figure S1**). Some cancer types ( _e.g._ pancreatic adenocarcinomas in the RAS
arm clusters 9 and 10, and breast cancers being excluded from the same cluster,
see **supplementary figure S1**) are unevenly distributed toward a specific
class, we could not identify cancer type that was exclusive to a single tumor
class. Somewhat surprisingly to us, this finding extends to histological
subtypes of breast, head and neck cancers, leukemias, _etc._ (see **Figure 1i**
and **Supplemental Figure S2**). Indeed, it would be surprising if we did not
observe bias for some cancers, as it has long been known that certain treatment
regimens are likely to work for a majority of patients of a given cancer type.
Nonetheless, the basis for some tumors being treatment-refractory in spite of
clinical indications remains elusive. Our analysis suggests the possibility of
more general mechanisms underlying tumor evolution regardless of tissue,
histological or cellular subtype and could form a basis for future separation of
patients into different precision treatment regimens.

@@

Finally, we must remark on two of the classes, 1 and 8, which for which we did
not find distinctive associations with pathways. Class 1 has a relatively low
level of mutated pathways overall, although it is broadly enriched in many of
the same tumor-promoting pathways common to the other groups. This likely does
not reflect a true difference in stages of tumor evolution, as our data show
clearly that this class is as likely to contain stage IV as it is to contain
stage I tumors. We speculate this class represents a group of tumors with large
aberrations in methylation, copy number or other genome rearrangements which we
excluded from the present analysis. In the future, we hope to incorporate these
other data into a pathway-centric analysis as we have done here for point
mutations and indels. Class 8 represents a group of hyper-mutated tumors that
harbor so many mutations that virtually no pathway is unaffected. This group
highlights one of the limitations of the approach that we’ve taken, in that it
is likely that a significant fraction of the samples flagged as “mutant” for
each pathway are actually unaffected, consistent with those samples have low
information content. Similar to class 1 we speculate that a future analysis of
other data types, including improved filtering of mutations, may resolve these
tumors into more meaningful categories.

@@

The fact that we are likely excluding many defects (under chromosomal
abberations, amplifications, demethylation and so on) also raises the
possibility that our classes are incomplete or may change at a future time when
we can account for them. For now at least, we think that the point mutation data
are representative of the most important pathways.

@@

Our findings imply that there are separate processes in the etiology of cancer
that can be broadly thought of as general cancer promoting, class-specific
mutations and metastasis. General cancer promoting processes must include
factors that relate to genome stability either directly or indirectly. Such
pathways are disrupted in all or most of the defined classes (see **figures 2**
and **supplementary figure S1**) and often are the result of abberations
involving common driver genes such as TP53, BRCA1/2, MMR genes, mitotic
checkpoints, Cohesion complexes, _etc._ (**supplementary figure S3** and
**supplemental file 1**). Class-specific evolution must involve the acquisition
of combinatorial disruptions to pathways that may individually be harmful but
together produce more specialized cancer phenotypes and increased fitness (as in
**figure 4** and **tables 1** and **ST2**). Importantly, our work does not
distinguish between the order in which these mutations should accumulate.
However, since many of the genes in the non-specific pathways involve genome
stability, it is reasonable to hypothesize that such mutations promote the
acquisition of class specific defects, and natural selection produces the
classes we observed.

>>>>

One of the biggest technical hurdles in tumor research is that the search space
comprising ~20,000 protein-coding genes is comparable with the number of tumor
samples, even with multiple mutations per sample
[@hornNetSigNetworkbasedDiscovery2018]. And that is not accounting for the fact
that ~30k non-coding genes and hundreds of thousands of non-coding regulatory
elements exist with oncogenic or tumor suppressor potential in spite of many
efforts detect non-random or selected mutation above the noise[@deesMuSiC2012;
@tamboreroOncodriveCLUST2013; @lawrenceDiscoverySaturationAnalysis2014;
@kumarStatisticallyIdentifyingTumor2015; @jiangWITER2019;
@zhaoModelingPositive2019; @non-coding-mutations-Zhang]. Hence, we sought to
investigate cancer as a disease of basic cellular and biochemical pathways,
which we accomplished by translating gene-level mutations into pathway level
disruptions.

The hypothesis that cancer is the result of dysfunction from a limited number of
basic cellular processes common to all living human cells was first
introduced and later expanded on in a pair of classic essays by Hanahan and
Weinberg [@weinberg200; @hallmarks_II]. In this view all tumors, regardless of
tissue of origin, acquire 6 essential phenotypes; sustaining proliferative
signaling, evading growth suppressors, replicative immortality, evading
apoptosis, supportive angiogenesis, and tissue invasion and metastasis
[@weinberg2000]. Advancements in the field over the ensuing decade led to the
inclusion of new tumor promoting processes, namely deregulation of celullar
energy cycle, tumor promoting inflammation, evasion of the immune system, genome
instability, and on top of these acquired traits a new appreciation for the
concept of the tumor microenvironment and its contribution to the evolution and
promotion of metastatic tumorigenesis [@hallmarks_II].

An alternative hypothesis is that every tumor belongs to one of a large number
of syndromes which are unique to each tissue-of-origin, that share some
mechanisms and treatment strategies. Recent publication of Encode, Roadmap and TCGA
consortium papers present a view largely consistent with the latter
hypothesis. Perhaps owing to the intractable complexity of genomic data in all
its forms, the inescapable conclusion thus far is that tissue-of-origin remains
the most important driver of tumor characteristics at every scale and by every
technique. Our observations contrast with this view, and instead support an
interpretation of publicly available  data in which all tumors manifest one of a
limited number of metastable phenotypes resulting from disruptions in basic
pathways represented by our classes (**figure 1a-k**), with increased fitness
regardless of tissue of origin, which then converge on end stage cancer by
accumulating complementary mutations to the other classes (**figure 1l** and
**Table 2**).

We find that most tumors segregate into one of 12 biologically distinct
clusters, regardless of the cancer type (**figure 1h** and **supplementary
figure S1**). Some cancer types ( _e.g._ pancreatic adenocarcinomas in the RAS
arm clusters 9 and 10, and breast cancers being excluded from the same cluster,
see **supplementary figure S1**) are unevenly distributed, though we could not
identify cancer type that was exclusive to a single tumor class. Somewhat
surprisingly to us, this finding extends to histological subtypes of breast,
head and neck cancers, leukemias, _etc._ (see **Figure 1i** and **Supplemental
Figure S2**). Indeed, it would be surprising if we did not observe bias for some
cancers, as it has long been known that certain treatment regimens are likely to
work for a significant fraction of patients of a given cancer type. Nonetheless,
the basis for some tumors being treatment-refractory in spite of clinical
indications remains elusive. Our analysis suggests the possibility of
identifying more informative molecular, histological or cellular subtypes and
could form a basis for future separation of patients into different precision
treatment regimens.

@@

@@

Finally, we must remark on the limitations of our work exemplified in two of the
classes, 1 and 8, which for which we did not find distinctive associations with
pathways. Class 1 has a relatively low level of mutated pathways overall,
although it is broadly enriched in many of the same tumor-promoting pathways
common to the other groups. This likely does not reflect a true difference in
stages of tumor evolution, as our data show clearly that this class is as likely
to contain stage IV metastatic tumors as it is to contain those of stage I. We
speculate this class represents a group of tumors with large aberrations in
methylation, copy number or other genome rearrangements which we excluded from
the present analysis. In the future, we hope to incorporate these other data
into a pathway-centric analysis as we have done here for point mutations and
indels. Now that we have established this as a feasible and potentially fruitful
approach to understanding the cancer genome, it may be worth expanded effort in
this area. In addition, Class 8 represents a group of hyper-mutated tumors that
harbor so many mutations that virtually no pathway is unaffected. It seems
likely that a significant fraction of the samples flagged as “mutant” for each
pathway are actually unaffected, consistent with those samples have low
information content. Similar to class 1 we speculate that a future analysis of
other data types, including improved filtering of mutations across all classes,
may resolve these tumors into improved classifications.

@@

Our findings imply that there are separate processes in the etiology of cancer
that can be broadly thought of as general cancer promoting, class-specific
mutations and metastasis. General cancer promoting processes must include
factors that relate to genome stability and immortality, as "enabling
characteristics" of the cancer phenotype [@hallmarks_II]. Such pathways are
disrupted in all or most of the defined classes (see **figures 2** and
**supplementary figure S1**) and often are the result of abberations involving
common driver genes such as BRCA1/2, MMR genes, mitotic checkpoints, Cohesion
complexes, _etc._ (**supplementary figure S3** and **supplemental file 1**).
Class-specific evolution must involve the acquisition of combinatorial
disruptions to pathways that may individually be harmful but together produce
more specialized cancer phenotypes and increased fitness (as in **figure 4** and
**tables 1** and **ST2**). Importantly, our work does not distinguish between
the order in which these mutations should accumulate. However, since many of the
genes in the non-specific pathways involve genome stability, it is reasonable to
hypothesize that such mutations promote the acquisition of class specific
defects, and natural selection produces the classes we observed.

One of the drawbacks of bulk tumor whole genome sequencing data is the problem
of tumor heterogeneity, both in terms of evolving subclones with different or
passively diverging mutation profiles which then serve as a substrate for futher
selection. Moreover consortium samples are likely to contain contamination from
support tissue and inflammatory and immune cells of the innate and adaptive
immune systems, all potentially harboring cancer supporting mutations
[@bhowmick_review]. We think it will be instructive to explore these ideas in
the context of single cell experiments.
